---
title: "Effect size estimation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Effect size for paired binary outcomes: Cohen's g}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: bibliography.bib  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Overview

The main purpose for using `mcnemar_table()` is to provide the user with the ability to calculate an effect size for each comparison of a treatment a control. Cohen's *g* is used for effect size [@cohen2013] and has been used in several studies with the McNemar test [@macdonald2022; @he2023; @goto2023; @kloosterman2024; @langer2024; @binkowska2026]. `mcnemar_table()` should be used only after using `mcnemar_control()`. In the example below, we fit the asymptotic McNemar test to our example `dat`, which has a control and three treatments.

```{r}
# load package
library(mcnemarControl)
# example data
dat <- mcnemar_example_long(seed=2)
# The asymptotic McNemar test with default p-value adjust using the Holm method 
dat.asymp <- mcnemar_control(dat, id = "id", condition = "condition", outcome = "outcome",
                               control = "Control", method="asymptotic",p_adjust="holm", ci=0.95)
```

## Cohen's *g*

Cohen's *g* has been used as an effect size with the McNemar test. It was designed to test if there was a significant departure of a population proportion from 0.5 and has been suggest as an effect size for the sign test [@cohen1992]. The sign test (look Siegal for definition.is similar to the McNemar test.

$$
g = P -0.5
$$

For the McNemar test, Cohen's *g* is a measure of the degree of asymmetry for the discordant pairs from a $2 \times 2$ contingency table, where

$$
g = {\frac{b}{b+c}} -0.5
$$

When the option `include_g = TRUE` is included in `mcnemar_table()`, the function calls [cohens_g](https://easystats.github.io/effectsize/reference/cohens_g.html) function from the **effectsize** package.

```{r}
dat.cohens_g.marginal <- mcnemar_table(dat.asymp, data=dat,include_g = TRUE)
dat.cohens_g.marginal
```

The output from `mcnemar_control()` shows a table with a row for each treatment comparisons with the control. The cells from the $2 \times 2$ contingency is presented by four columns, with the discordant cells represented by columns `b` and `c`, and the concordant cells by `a` and `d`. Finally, column cohens_g shows Cohen's *g* that was estimated from the [cohens_g](https://easystats.github.io/effectsize/reference/cohens_g.html) function. No confidence intervals are shown, as the `methodCI = "none"`which is the default has been used here. `mcnemar_control()` includes a marginal and two methods for adjusting the estimates for the confidence intervals to account for multiple McNemar test comparisons.

### Marginal confidence intervals {#sec-marginalCI}

When `method="marginal"` is included with or without `ci_level = 0.95`, 95% confidence intervals are included in the output table. The marginal confidence intervals are **single-compariso**n $\text{CI}$ from the [cohens_g](https://easystats.github.io/effectsize/reference/cohens_g.html) function that were suppressed with `methodCI = "none"`.

```{r}
dat.cohens_g.marginal <- mcnemar_table(dat.asymp, data=dat,include_g = TRUE,method="marginal", ci_level = 0.95)
dat.cohens_g.marginal

```

### Simultaneous max-\|t\| confidence intervals {#sec-simultaneousCI}

The `method="simultaneous"` provides the user with simultaneous confidence intervals for Cohen’s $g$ were obtained using a **subject-level bootstrap max-\|*t*\|** procedure, where bootstrap replicates of the vector of treatment-specific $g$ estimates are studentized, the maximum absolute studentized deviation over treatments are recorded in each replicate, and the resulting bootstrap distribution is used to derive a **single-step family-wise critical value** for simultaneous confidence intervals.This procedure is analogous to other bootstrap max-*t* simultaneous-inference methods [@kuchibhotla2022], such as the [simultaneous_ci](https://rdrr.io/cran/PosiR/man/simultaneous_ci.html) function in the **PosiR** package[@Chukwuma2025], but is specialized to paired binary McNemar-type effect sizes rather than regression coefficients across a model universe. It consists of the following steps:

1.  Re-sample subjects with replacement;

2.  Recompute the full vector of treatment-specific estimates for Cohen's *g*;

3.  Studentize the bootstrap deviations;

4.  Take the **maximum absolute studentized deviation** across the treatment family in each bootstrap replicate;

5.  Use the bootstrap distribution of that maximum to get one common critical value,

6.  Build simultaneous intervals for every treatment using that family-wise critical value.

```{r}
dat.cohens_g.simultaneous <- mcnemar_table(dat.asymp, data=dat,include_g = TRUE,method="simultaneous", ci_level = 0.95)
dat.cohens_g.simultaneous
```

#### Step-by-step procedure for simultaneous max-\|t\| confidence intervals on g scale {#sec-maxTg}

The procedure can be summarised as follows:

1.  Reconstruct subject-level wide data

    The function starts with the original long-format repeated data and reshapes them into **one row per subject**, with one column for the control outcome and one column for each treatment outcome. It keeps only complete subject profiles for the simultaneous calculation.

2.  Recode outcomes to binary values

    Using `outcome_levels`, outcomes are recoded so that the **second level** is treated as success (`1`) and the first as failure (`0`). This binary representation is allows the discordant counts $b$ and $c$ to be reconstructed in each comparison and in each bootstrap sample.

3.  Compute the original vector of Cohen's *g* estimates

    For each treatment $j$, the code compares treatment with control on the same subjects and computes

    $$
    \hat{g} = \frac{b_{j}}{b_{j}+c_{j}}-0.5
    $$

    when $b_{j}+c_{j}>0$; otherwise it returns $\text{NA}$. These original values form the vector of observed effect sizes across the treatment family.

4.  Bootstrap the subjects

    The code then re-samples the subjects with replacement $R$ times, carrying along each subject’s full repeated outcome profile across control and all treatments. This preserves the correlation across treatment-vs-control comparisons in the re-sampled data.

5.  Recompute the vector of Cohen's *g* values in each bootstrap sample

    For each bootstrap replicate, the statistic function recomputes the entire treatment-specific vector

    $$
    \hat{g}_{1}^*,
    \hat{g}_{2}^*,...
    \hat{g}_{K}^*
    $$

    These bootstrap vectors are stored in the bootstrap matrix `Tmat`.

6.  Estimate a bootstrap standard error for each treatment effect

    From the bootstrap matrix, the code computes the bootstrap standard deviation ($\text{sd}$) for each treatment-specific Cohen's g value:

    $$
    \hat{\text{SE}} = \text{sd}(\hat{g}_{j}^{*(1)},...\hat{g}_{j}^{*(R)})
    $$

    These treatment-specific standard errors ($\text{SE}$s) are then used to studentize the bootstrap deviations.

7.  Keep only usable comparisons / replicates

    The code identifies which treatment comparisons have finite original estimates and positive finite bootstrap $\text{SE}$s. It also identifies which bootstrap rows have all required entries finite for the family-wise calibration. Treatments with zero or undefined $\text{SE}$s are flagged as **degenerate** and users are advised to use the marginal $\text{CI}$s.

8.  Form studentized bootstrap deviations

    For each usable treatment $j$ and each usable bootstrap replicate $r$, the code computes the standardized deviation

    $$
    Z_{j}^{*(r)} = \frac{\hat{g_{j}}^{*(r)}-\hat{g_{j}}}{\hat{\text{SE}_{j}}}
    $$

    This is the "*t*" part of the max*-t* method with each treatment’s bootstrap fluctuation measured with its own standard-error units.

9.  Compute the $\text{max}-|t|$ statistic for each bootstrap replicate

    For each bootstrap replicate replicate $r$, the method takes the **largest absolute** standardized deviation across the treatment family:

    $$
    M^{*(r)}= \max_{j} |Z_{j}^{*(r)}|
    $$

    This is the family-wise reference statistic. It captures the worst standardized error anywhere in the treatment family for that bootstrap sample.

10. Extract the simultaneous critical value

    The code then takes the empirical $(1-\alpha)-$quantile of the bootstrap `maxabs` values to get the simultaneous critical value $c_{\alpha}$, which is used for every treatment-specific interval.

11. Build the simultaneous $\text{max}-|t|$ confidence intervals

    Finally, for each treatment $j$, the code constructs the interval

    $$
    \hat{g_{j}}\pm c_{\alpha}\hat{\text{SE}_j}
    $$

    These are returned as `ci_low` and `ci_high` for the simultaneous max-*t* $\text{CI}$s.

12. Flag degenerate comparisons

    If a comparison has no discordant pairs, or if its bootstrap $\text{SE}$ is zero / non-finite, the method flags it as **degenerate** and the interval is not treated as a regular simultaneous $\text{CI}$. A degeneracy **flag** is explicitly stored for later uses and marginal intervals are recommended instead.

#### Simultaneous max-\|t\| logit confidence intervals {#sec-maxTlogit}

One of the disadvantages of calculating the confidence interval directly on from Cohen's *g* is that the simultaneous confidence intervals can have estimated values above the range limit of $[-0.5, 0.5]$.

In order to obtained confidence interval estimates within the limits, we can use `methodCI = "simultaneous_logit"` that then allows `mcnemar_table()` to compute the observed discordant-pair proportion $p = b/(b+c)$ for each treatment, transforms it to $\eta = \text{logit}()p$, then bootstraps subjects to obtain a family of bootstrap $\eta^{*}$ values, forms the maximum absolute standardized deviation over treatments in each bootstrap replicate, and uses the $(1-\alpha)-$quantile of that $\text{max}-|t|$ distribution as a simultaneous critical value, and then back-transforms the resulting logit-scale limits to $g = p-0.5$. The resulting interval endpoints are therefore naturally bounded within $[−0.5,0.5]$ whenever defined.

```{r}
dat.cohens_g.simultaneous_logit <- mcnemar_table(dat.asymp, data=dat,include_g = TRUE,method="simultaneous_logit", ci_level = 0.95)
dat.cohens_g.simultaneous_logit
```

#### Step by step procedure for simultaneous max-\|t\| confidence intervals on logit scale

The first two steps are the same as for the procedure on the original *g* scale @sec-maxTg. The procedure then continues from:

3.  Compute the proportion *p* of observed discordant-pairs

    For each treatment $j$,

    $$
    \hat{p_{j}}= \frac{b+{j}}{b_{j}+c_{j}}
    $$

    whereby $b_{j}$ is the number of subjects where the control $=1$ and treatment $= 0$, and $c_{j}$ is the number of subjects where the control $=0$ and treatment $= 1$ and provided $b_{j}+ c_{j} \gt 0$. If there are no discordant pairs, the estimate is treated as undefined ($\text{NA}$).

4.  Transform $p$ to the logit scale

    For each treatment,

    $$
    \hat{\eta_{j}}= \text{logit}(\hat{p_{j}}) = \log \bigg(\frac{\hat{p_{j}}}{1-\hat{p_{j}}}\bigg)
    $$

    but only when $0 \lt \hat{p_{j}} \lt 1$. $\hat{p_j}$. if $\hat{p_j} = 1$, $\hat{p_j} = 0$ or are undefined, they are marked as **degenerate on the observed scale** and excluded from the $\text{max}-|t|$ calibration.

5.  Generate bootstrap replicates on the logit scale

    The function then defines a bootstrap statistic that, for each subject-level bootstrap sample:

    1.  reconstructs the treatment-versus-control discordant counts $b_{j}$ and $c_{j}$

    2.  computes

        $$
        p_{j}^{*} = \frac{b_{j}^{*}}{b_{j}^{*}+c_{j}^{*}}
        $$

        provided there are discordant pairs,

    3.  then transforms to

        $$
        \eta_{j}^{*} = \log{\frac{p_{j}^{*}}{1-p_{j}^{*}}}
        $$

        only if $0 \lt p_{j}^{*} \lt 1$. Otherwise $\text{NA}$. This is done using the `boot::boot()` function on the **subject rows** of the wide repeated-measures dataset. This is importan as it preserves the dependence structure across all treatment-versus-control comparisons because each bootstrap replicate carries the **full repeated outcome profile** of each resampled subject. This is what makes the intervals simultaneous at the **family level** across treatments.

6.  Estimate bootstrap standard errors on the logit scale

    From the matrix of bootstrap logit estimates, the function computes a treatment-specific bootstrap standard deviation:

    $$
    \hat{\text{SE}_{n,j}}= \text{sd}(\eta_{j}^{*})
    $$

    using a finite bootstrap replicates only. Treatments are considered usable and are retained for $\text{max}-|t|$ calibration only if:

    - the observed treatment $\eta_{j}^{*}$ is **not degenerate**,

    - the bootstrap $\hat{\text{SE}_{n,j}}$ is finite,

    - and the bootstrap $\hat{\text{SE}_{n,j}} > 0$.

7.  Build the $\text{max}-|t|$ bootstrap statistics

    For each bootstrap replicate and each usable treatment, the function computes the standardized deviation:

    $$
    Z_{j}^{*} = \frac{\eta_{j}^{*}-\hat{\eta_{j}}}{\hat{\text{SE}_{n,j}}}
    $$

    Then, for each replicate, it computes the familywise maximum statistic:

    $$
    M^{*(r)}= \max_{j} |Z_{j}^{*(r)}|
    $$

8.  Extract the simultaneous critical values

    The code then takes the empirical $(1-\alpha)-$quantile of the bootstrap `maxabs` values to get the simultaneous critical value $c_{\alpha}$, which is used for every treatment-specific interval.

9.  Construct simultaneous intervals on the logit scale

    For each usable treatment, the function determines simultaneous confidence limits on the transformed scale:

    $$
    \hat{\eta_{j}}\pm c_{\alpha}\hat{\text{SE}_{n,j}}
    $$

10. Back-transform to the *g*-scale

    The transformed limits are back-converted to the probability scale:

    $p_{L,j} = \text{logit}^{-1}(\eta_{L,j})$ and $p_{U,j} = \text{logit}^{-1}(\eta_{U,j})$

    and then shifted back to Cohen’s *g*:

    $g_{L,j}= p_{L,j}-0.5$ and $g_{U,j}= p_{U,j}-0.5$

    So instead of smoothing or extrapolating beyond the boundary $[-0.5,0.5]$, the interval is **collapsed to the natural endpoint**. This is especially useful as the intervals respects the logical parameter range. i.e. Th confidence interval will be within $0.5$ at the upper limit.

11. Flag degenerate treatments

    Finally, the method marks treatments as degenerate if they were boundary/no-discordant cases or if their bootstrap $\text{SE}$ on the logit scale is zero or non-finite. These flags are returned in the output so downstream print / summary methods can recommend marginal intervals when appropriate.

## Difference

d

# References {#sec-ref_effectsize}
