---
title: "Introduction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get Started}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

`mcnemarControl` provides a control-vs-treatment McNemar test
comparisons.

## Data requirements

Your input data must be in **long format** with at least:

-   an `ID` column identifying paired observations (e.g., subject),
-   a `condition` column containing the control and one or more
    treatments,
-   a `binary` outcome column.

For each `id`, there must be **at most one row per condition**; if you
have duplicates (multiple rows per `id`–`condition`), you must aggregate
or resolve them before analysis.

In the implementation used by `mcnemar_control()`, data are reshaped to
wide format (one row per `id`) and then paired outcomes are compared
between control and each treatment.

> Practical tip: explicitly set `outcome_levels` (see Interpretation
> notes) to avoid ambiguity in directionality.

## Basic example

The example data in this package’s is a long-format example `dat`

```{r setup}
library(mcnemarControl)
```

The `mcnemar_control()` function uses the long formatted table in `dat`
to create a 2x2 contingency table (this is `tab` in the function). This
is built for each control–treatment pair from the paired outcomes for
the same subjects (`id`). The key idea is:

-   Rows = the control outcomes (`x`)
-   Columns = the treatment outcomes (`y`)
-   The row/column order is determined by outcome_levels (either
    supplied by you, or inferred)

The order of the outcome levels determines what is contained in each
cell of the contingency table. By default the outcome_levels =
c("No","Yes"). This then allows the part of the function `run_mcmenar()`
to calculate the number of discordant values (i.e. `b+c` from the
contigency table)

```{r}
dat <- mcnemar_example_long(n = 40, seed = 1)

fit <- mcnemar_control(
  dat,
  id = "id",
  condition = "condition",
  outcome = "outcome",
  control = "Control"
)
fit
```

The output includes a results table (one row per treatment comparison)
and the analysis settings used.

```{r}

summary(fit)
```

## Choose method (asymptotic / cc / midp)

`mcnemar_control()` supports three methods:

-   `"asymptotic"`: the standard asymptotic McNemar test,
-   `"cc"`: asymptotic McNemar test with continuity correction,
-   `"midp"`: McNemar mid-P approach.

> Note: `midp` does not return a Z statistic (it is reported as `NA`),
> but it returns p-values.

### method = "asymptotic"

`contingencytables::McNemar_asymptotic_test_paired_2x2()` uses the
uncorrected statistic

$$
Z = \frac{b-c}{\sqrt{(b + c)}}
$$ The test statistic `Z` can be negative or positive (Signed),
depending on whether $b < c$ \< or $b > c$.

-   If $b < c$, then $b−c < 0 ⇒ Z < 0$
-   If $b >c$\`, then $b−c > 0 ⇒ Z > 0$

For the asymptotic method, the sign of `Z` reflects the direction of the
imbalance between discordant cells `b` and `c` (given your chosen
outcome_levels order).

### method = "cc": Z is non-negative by design

`contingencytables::McNemar_asymptotic_test_CC_paired_2x2()` uses the
continuity-corrected statistic

$$
Z_{cc} = \frac{\vert{b-c}\rvert-1}{\sqrt{(b + c)}}
$$

Because it uses $\vert{b-c}\rvert$, the returned $z_{cc}$ is always
$≥ 0$

### Example: compare methods

```{r}
fit_asym <- mcnemar_control(
  dat, "id", "condition", "outcome",
  control = "Control",
  method = "asymptotic"
)

fit_cc <- mcnemar_control(
  dat, "id", "condition", "outcome",
  control = "Control",
  method = "cc"
)

fit_midp <- mcnemar_control(
  dat, "id", "condition", "outcome",
  control = "Control",
  method = "midp"
)

summary(fit_asym)
summary(fit_cc)
summary(fit_midp)

```

## Multiple comparisons (Holm default, BH/Hochberg options)

By default, `mcnemar_control()` uses **Holm** adjustment for family-wise
error rate control. You can switch to:

-   `"BH"` for false discovery rate control,
-   `"hochberg"` for family-wise error rate control under certain
    dependence assumptions

```{r}
fit_holm <- mcnemar_control(
  dat, "id", "condition", "outcome",
  control = "Control",
  method = "asymptotic",
  p_adjust = "holm"
)

fit_bh <- mcnemar_control(
  dat, "id", "condition", "outcome",
  control = "Control",
  method = "asymptotic",
  p_adjust = "BH"
)

fit_hoch <- mcnemar_control(
  dat, "id", "condition", "outcome",
  control = "Control",
  method = "asymptotic",
  p_adjust = "hochberg"
)

summary(fit_holm)
summary(fit_bh)
summary(fit_hoch)
```

The results table includes both raw p-values and adjusted p-values.

## Effect sizes (Cohen’s g and CI)

`mcnemar_control()` reports Cohen’s *g* for paired 2×2 tables, along
with a confidence interval (CI level set by `ci`).

### Example: change CI level

```{r}
fit_ci90 <- mcnemar_control(
  dat, "id", "condition", "outcome",
  control = "Control",
  method = "asymptotic",
  ci = 0.90
)

fit_ci95 <- mcnemar_control(
  dat, "id", "condition", "outcome",
  control = "Control",
  method = "asymptotic",
  ci = 0.95
)

summary(fit_ci90)
summary(fit_ci95)
```

## Interpretation notes

### Interpreting the McNemar Z statistic across methods

In `mcnemar_control()`, the reported McNemar statistic `Z` comes from
the underlying **contingencytables** functions. For
`method = "asymptotic"`, `Z` is signed and can be negative or positive
because it is computed from the difference of discordant counts (e.g.,
proportional to $b - c$). For `method = "cc"`, the continuity-corrected
statistic uses $|b - c|$ (so the returned `Z` is non-negative) and
should be interpreted as a magnitude-only statistic. In both cases, the
default p-value is two-sided and is computed from $|Z|$, so the sign
does not affect the p-value.

### (1) `Z` sign conventions: asymptotic vs cc

-   With `method = "asymptotic"`, the reported `Z` can be **negative or
    positive** (signed statistic).
-   With `method = "cc"`, the reported `Z` is **non-negative**
    (magnitude-only) because the continuity-corrected form uses an
    absolute difference internally.

This is expected and does not indicate an error; two-sided p-values
depend on `abs(Z)`.

### (2) Default `outcome_levels` ordering

If you do not specify `outcome_levels`, the function chooses an order
based on the **first appearance** of outcome values in your data after
removing missing values. This can vary by dataset and can change the
sign of the asymptotic `Z`.

**Recommendation:** explicitly specify outcome ordering so the table
layout and directionality are stable:

```{r, eval=FALSE}
fit <- mcnemar_control(
  dat,
  id = "id",
  condition = "condition",
  outcome = "outcome",
  control = "Control",
  outcome_levels = c("No", "Yes")
)
```

This ensures: - `tab[1,2]` corresponds to control=`"No"`,
treatment=`"Yes"`, - `tab[2,1]` corresponds to control=`"Yes"`,
treatment=`"No"`, and therefore the meaning of discordant counts and the
sign of `Z` are interpretable.

### Contigency table functionality

A `mcnemar_table()` has been added to provide a table that summarises
the values for each contingency table with some additional
functionality. By dafault `mcnemar_table()` produces a table with only 4
columns, each representing one of the cells from the contingency table.
(i.e. in the order `a`, `b`, `c`, `d`.). default can then be used with
the `DiscreteTests` and `DiscreteFWER` package. This allows the user to
use the `DHolm()` function to calculate p-values for the mcnemar tests
functions from the `DiscreteTests` package. The `mcnemar_test_pv()`
function can perform the asymptotic or exact Mcnemar test.
